Minimal trap spaces of Boolean networks are maximal siphons of their Petri net encoding

### 2022-12-22 manuscripts/TCS2022/tcs2022.tex
Petri net definition

\end{definition}
A \emph{marking} for a Petri net is a mapping \(m : P \mapsto \mathbb{N}\) that assigns a number of tokens to each place. A place \(p\) is marked by a marking \(m\) if and only if \(m(p) > 0\). We shall write \(pred(x)\) (resp.\ \(succ(x)\)) to represent the set of vertices that have a (non-zero weighted) arc leading to (resp.\ coming from) \(x\).

The link between Boolean networks \emph{à la} Thomas and Petri nets was originally established in~\cite{chaouiya2004qualitative} in order to make available formal methods like model-checking for the analysis of such systems.
The basic encoding into 1-safe (i.e., never more than one token in each place) nets only holds for purely Boolean networks but was later extended to multivalued Logical models in two ways, either in~\cite{chaouiya2011petri} with non 1-safe Petri nets or more recently in~\cite{chatain2014characterization} with 1-safe nets but many more places.

Since our study is focused on Boolean networks, we briefly recall the original encoding here.
Its basis is that every node (\emph{gene}) \(v\) of the original model \(\mathcal{M} = (V, F)\) is represented by two separate places (\(p_v\) and \(\overline{p}_v\)), corresponding to its two states, active, and inactive, respectively.
Each conjunct of the logical function that activates the \emph{gene} will lead to a transition \(t\), consuming the inactive place (i.e., a directional arc from \(\overline{p}_v\) to \(t\)), producing the active place (i.e., a directional arc from \(t\) to \(p_v\)), and with all other literals both consumed and produced (i.e., a bidirectional arc).
And conversely for the inactivation.
Let \(s\) be a state of the Boolean network and \(m_s\) be its corresponding marking in the encoded Petri net. It holds that \(\forall v \in V\), \(s(v) = 0\) if and only if \(m_s(\overline{p}_v) = 1\) and \(s(v) = 1\) if and only if \(m_s(p_v) = 1\). Note also that at any marking \(m\) of the Petri net encoding a Boolean network, it always holds that \(m(p_v) + m(\overline{p}_v) = 1\).

The main property of this encoding is that it is completely faithful with respect to the update scheme of the original Boolean network.
For each node \(v\) of \(\mathcal{M}\), only transitions corresponding to \(v\) can change the current marking of \(p_v\) or \(\overline{p_v}\).
In addition, at any marking at most one of such transitions is enabled because \(m(p_v) + m(\overline{p}_v) = 1\) holds.
Hence, for any update scheme in \(\mathcal{M}\), we have a corresponding firing scheme in \(\mathcal{P}\), which preserves the equivalence between the dynamics of \(\mathcal{M}\) and \(\mathcal{P}\)~\cite{DBLP:journals/nc/ChatainHKPT20}.

For illustration, let us reconsider the Boolean network shown in Example~\ref{example:BN}.
Figure~\ref{fig:PN} shows the Petri net encoding of this Boolean network.
Place \(p_{x_1}\) (resp. \(\overline{p_{x_1}}\)) in \(\mathcal{P}\) represents the activation (resp.\ the inactivation) of node \(x_1\) in \(\mathcal{M}\).
Marking \(\{p_{x_1}, \overline{p_{x_2}}\}\) in \(\mathcal{P}\) represents state 10 in \(\mathcal{M}\).
Transitions \(t^{1}_{x_1}\) and \(t^{2}_{x_1}\) represent the update of node \(x_1\).
Of course, in any marking \(t^{1}_{x_1}\) and \(t^{2}_{x_1}\) cannot be both enabled.
Then, the fully asynchronous update scheme in \(\mathcal{M}\) corresponds to the classical firing scheme in \(\mathcal{P}\) where only one of the enabled transitions for a given marking will be fired~\cite{Murata1989}.

>>>>

\end{definition}
A \emph{marking} for a Petri net is a mapping \(m : P \mapsto \mathbb{N}\) that assigns a number of tokens to each place. 
A place \(p\) is marked by a marking \(m\) if and only if \(m(p) > 0\). 
Marking \(m\) can be seen as a subset of \(P\) that contains all marked places by \(m\).
We shall write \(pred(x)\) (resp.\ \(succ(x)\)) to represent the set of vertices that have a (non-zero weighted) arc leading to (resp.\ coming from) \(x\).
In this work, we consider a class of Petri nets called 1-safe Petri nets where every place has at most 1 token and all arcs are of weight 1.
In this case, weights are implicitly omitted in the arcs of a Petri net.
Then, a transition \(t \in T\) is \emph{enabled} at a marking \(m\) if and only if \(pred(t) \subseteq m\). 
The firing of \(t\) leads to a new marking \(m'\) specified by \(m' = (m \backslash pred(t)) \cup succ(t)\).
Note that when multiple transitions are enabled, we need to embed one firing scheme (similar to the update scheme of a Boolean network) to the Petri net.
The classical firing scheme is that only one of the enabled transition is non-deterministically chosen to fire~\cite{Murata1989}.

The link between Boolean networks \emph{à la} Thomas and Petri nets was originally established in~\cite{chaouiya2004qualitative} in order to make available formal methods like model-checking for the analysis of such systems.
The basic encoding into 1-safe (i.e., never more than one token in each place) nets only holds for purely Boolean networks but was later extended to multivalued logical models in two ways, either in~\cite{chaouiya2011petri} with non 1-safe Petri nets or more recently in~\cite{chatain2014characterization} with 1-safe nets but many more places.

Since our study is focused on Boolean networks, we briefly recall the original encoding here.
Its basis is that every node (\emph{gene}) \(v\) of the original model \(\mathcal{N} = (V, F)\) is represented by two separate places (\(p_v\) and \(\overline{p}_v\)), corresponding to its two states, active, and inactive, respectively.
Each conjunct of the logical function that activates the \emph{gene} will lead to a transition \(t\), consuming the inactive place (i.e., a directional arc from \(\overline{p}_v\) to \(t\)), producing the active place (i.e., a directional arc from \(t\) to \(p_v\)), and with all other literals both consumed and produced (i.e., a bidirectional arc).
And conversely for the inactivation.
Let \(s\) be a state of the Boolean network and \(m_s\) be its corresponding marking in the encoded Petri net. 
It holds that \(\forall v \in V\), \(s(v) = 0\) if and only if \(m_s(\overline{p}_v) = 1\) and \(s(v) = 1\) if and only if \(m_s(p_v) = 1\). Note also that at any marking \(m\) of the Petri net encoding a Boolean network, it always holds that \(m(p_v) + m(\overline{p}_v) = 1\).

The main property of this encoding is that it is completely faithful with respect to the update scheme of the original Boolean network.
For each node \(v\) of \(\mathcal{N}\), only transitions corresponding to \(v\) can change the current marking of \(p_v\) or \(\overline{p_v}\).
In addition, at any marking at most one of such transitions is enabled because \(m(p_v) + m(\overline{p}_v) = 1\) holds.
Hence, for any update scheme in \(\mathcal{N}\), we have a corresponding firing scheme in \(\mathcal{P}\), which preserves the equivalence between the dynamics of \(\mathcal{N}\) and \(\mathcal{P}\)~\cite{DBLP:journals/nc/ChatainHKPT20}.

For illustration, let us reconsider the Boolean network shown in Example~\ref{example:BN}.
Figure~\ref{fig:PN} shows the Petri net encoding of this Boolean network.
Place \(p_{x_1}\) (resp. \(\overline{p_{x_1}}\)) in \(\mathcal{P}\) represents the activation (resp.\ the inactivation) of node \(x_1\) in \(\mathcal{N}\).
Marking \(\{p_{x_1}, \overline{p_{x_2}}\}\) in \(\mathcal{P}\) represents state 10 in \(\mathcal{N}\).
Transitions \(t^{1}_{x_1}\) and \(t^{2}_{x_1}\) represent the update of node \(x_1\).
Of course, in any marking \(t^{1}_{x_1}\) and \(t^{2}_{x_1}\) cannot be both enabled.
Then, the fully asynchronous update scheme in \(\mathcal{N}\) corresponds to the classical firing scheme in \(\mathcal{P}\) where only one of the enabled transitions for a given marking will be fired~\cite{Murata1989}.
